**Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces**

### 2025-06-21 paper/manuscript.qmd
update based on comments

---
title: "**Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces**"
shorttitle: "Choice-Confidence (CHOCO) Model"
author:
  - name: Dominique Makowski
    corresponding: true
    orcid: 0000-0001-5375-9967
    email: D.Makowski@sussex.ac.uk
    url: https://realitybending.github.io/
    roles:
      - Conceptualization
      - Data curation
      - Formal Analysis
      - Funding acquisition
      - Investigation
      - Methodology
      - Project administration
      - Resources
      - Software
      - Supervision
      - Validation
      - Visualization
      - Writing – original draft
    affiliations:
      - id: "id1"
        name: "University of Sussex"
        department: "School of Psychology"
        city: "Brighton"
        country: "UK"
        postal-code: "BN1 9RH"
      - name: "University of Sussex"
        department:  "Sussex Centre for Consciousness Science"
  - name: Ana Neves  
    orcid: 0009-0006-0020-7599
    role:
      - Data curation
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - ref: "id1"
  - name: Andy Field 
    orcid: ""
    role:
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - ref: "id1"
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    gratitude: |
      ::: {.callout-note icon=false appearance="simple"}
      This preprint is a non-peer-reviewed work from the [**Reality Bending Lab**](https://realitybending.github.io/).
      ![](https://realitybending.github.io/media/ReBeL_LogoOnly_hu11484441381606756729.png){width=20% fig-align="center"}
      :::
    study-registration: ~
    data-sharing: ~
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    authorship-agreements: ~
abstract: |
  As AI-generated content becomes increasingly indistinguishable from real stimuli, understanding how individuals form beliefs about the reality of what they perceive is both theoretically and practically urgent. This study introduces the Choice-Confidence (CHOCO) model as a tool to dissect the cognitive mechanisms underlying decision-making judgments. Conceptualized as a mixture of two Beta distributions, CHOCO simultaneously estimates the probability of selecting one category (e.g., "real" vs. "AI-generated") and the confidence associated with each choice. We apply this model to two datasets (N=141 and N=189) in which participants judged whether faces were real photographs or AI-generated images, finding that facial attractiveness systematically increased the likelihood of being judged as "real", particularly for male participants. We discuss how these findings relate to the broader effect of facial attractiveness, and suggest future directions for both the cognitive and psychometric application of the CHOCO framework.
#   **Significance Statement**. TODO.
keywords: "subjective scales, choice confidence model, reality beliefs, AI-generated faces, attractiveness, simulation monitoring"
# Add more keywords if possible: beauty, Ordered Beta model, Bimodal data, decision-making
numbered-lines: true
floatsintext: true
mask: false
bibliography: bibliography.bib
format:
  apaquarto-pdf:
    documentmode: jou
    keep-tex: true
  apaquarto-docx: default
---

@@

<!-- Problem statement -->
Despite significant advancements in psychological science following the replication crisis [@open2015estimating], its progress is still hindered by its sub-optimal (or inappropriate) usage of statistical tools [@makowski2023we; @cumming2014new; @blanca2018current]. 
A prevalent issue is the continued reliance on linear models that assume normally distributed (Gaussian) data^[More specifically, that the outcome is distributed according to a Normal distribution which parameters are expressed as a linear function of the predictors.] - as this assumption often does not hold true for many types of psychological outcomes.
For instance, reaction times typically exhibit skewed distributions, choices can be represented as binary variables, and count data consists of strictly positive integers. 
Applying models that presume normality and model the "mean" of the outcome variable can lead to misinterpretation and potentially misleading conclusions when applied indiscriminately. It is thus important that psychologists use models that can best describe (or generate) the data they collect, to fully exploit them and bring more nuance and accuracy to their conclusions.

<!-- Bounded data, Beta-models and ZOIB -->
Among the most commonly collected data in psychology are responses on subjective scales, such as Likert-type items or visual analog scales, which exhibit some fundamental properties: these responses are bounded (and can be rescaled to a 0-1 range) and frequently display clustering at the extremes. 
Traditional linear models being ill-suited for such data, researchers have turned to using Beta distributions to model this data (instead of Gaussian), suited for continuous data within the (0,1) interval (i.e., excluding extreme responses). 
To address the frequent occurrence of exact zeros and ones (i.e., extreme values), zero-one inflated beta (ZOIB) models have been developed [@ospina2012general] to accommodate the excess of boundary values by incorporating additional components that model the probabilities of responses at 0 and 1 as a separate, independent process.

@@

The Beta-Gate model is a reparametrized Ordered Beta model [@kubinec2023ordered]^[In the Ordered Beta model, the cutpoints on the log-scale are directly used as parameters, instead of being derived from *pex* and *bex*.] available in the *cogmod* package in R (https://github.com/DominiqueMakowski/cogmod), in which participants' answers on bounded scales are conceptualized as latent responses that can fall past a pair of probabilistic "gates" (or cutpoints) that control whether the response is recorded as an extreme (0 or 1) or as a nuanced, continuous value in between (@fig-one). 
These distance of these gates from the edges of the scale varies based on two interpretable parameters: *pex* (the propensity by which people are likely to answer extreme values), and *bex*, a bias toward the upper extreme (1) versus the lower (0). 
A person's internal response that lies close to the edge might be "caught" by a gate and recorded as an extreme, while others pass through to express a continuous response (Beta-distributed with $\mu$ (*mu*) and $\phi$ (*phi*) as its mean and precision parameters).
The Beta-Gate model is based on the idea that extreme values can emerge not just from a fundamentally different underlying processes - as assumed in ZOIB models - but from a common process governed by thresholds of decisiveness and confidence.

```{r}
#| label: fig-one
#| fig-cap: "The Beta-Gate Distribution is a reparametrized ordered Beta model (Kubinec, 2023) that is governed by 4 parameters. 'Mu' and 'phi' correspond to the mean and precision of the continuous part of the distribution (between 0 and 1), and 'pex' (propensity of extremes) and 'bex' (balance of extremes) indirectly control the proportion of zeros and ones by specifying the location of the \"gates\", past which the latent response process is likely to generate extreme values. Specifically, 'pex' defines the total distance of both gates from the extremes (in yellow), and 'bex' determines the proportion of the right gate distance relative to the left. In this example, the total distance from the extremes is 'pex' = 0.1, with 40% ('bex' = 0.4) of that distance being on the right (and 60% on the left). The left gate is thus located at 0.6, and the right at 1-0.04 = 0.96."
#| apa-twocolumn: false  # A Figure Spanning Two Columns When in Journal Mode
#| out-width: "100%"

@@

- $P(x = 0) = \text{logistic}\left( \text{logit}(pex \cdot (1 - bex)) - \text{logit}(\mu) \right)$
- $P(x = 1) = 1 - \text{logistic}\left( \text{logit}(1 - pex \cdot bex) - \text{logit}(\mu) \right)$
- $P(x \in (0,1)) = 1 - P(x = 0) - P(x = 1)$

@@

```{r}
#| label: fig-two
#| fig-cap: "The CHOCO Model uses a mixture of Beta-Gate distributions to model separately the right and left sides of the scale (e.g., a rating of whether a statement was 'Truth' vs. 'Lie'), as well as their relative proportion. In this example, the participants are more likely overall (p = 60%) to select the right side of the scale ('Lie') than the left ('Truth'). They are also more confident in their choice (confright = 0.6 vs. confleft = 0.3). Extreme values (zeros and ones) are governed by the same mechanism as for Beta-Gate models."
#| apa-twocolumn: false  # A Figure Spanning Two Columns When in Journal Mode
#| out-width: "100%"

knitr::include_graphics("./figures/fig2.png")
```

@@

**Study 1** aims to evaluate the CHOCO model’s ability to better capture subjective scale responses that (potentially) reflect an underlying discrete choice, in comparison to existing models such as the ZOIB and Beta-Gate. Specifically, we will assess whether 1) CHOCO provides improved model fit, 2) yields deeper insights into population-level effects than traditional approaches (gender differences in reality beliefs), and 3) allows for the reliable estimation of interpretable individual-level parameters through random effects. **Study 2** will apply this model to more subtle effects, such as the effect perceived facial attractiveness on reality judgments, and test the ability to fit alternative data structures (scales with ordinal response options and mid-points). To this end, we analyze data from two separate studies in which participants judged whether a face image was AI-generated ("fake") or a real photograph.

@@

In today's post-truth era, the proliferation of advanced AI technologies has made it increasingly challenging to distinguish between authentic and synthetic media, bearing significant implications for information integrity and public trust [@lewandowsky2017beyond]. As traditional cues become less reliable (e.g., visual glitches and artefacts in generated images; formulaic generated text, etc.), people increasingly depend on contextual information and cognitive heuristics to assess authenticity, a process referred to as "simulation monitoring" [@makowski2019phenomenal].

@@

Using the open-access data from @makowski2025too, we included all heterosexual and bisexual (as these two groups did not seem to differ based on preliminary analyses and were thus grouped to maximize power) male and female participants, for a final sample of 141 participants (Mean age = 28.4, SD = 9.0, range: [19, 66]; Sex: 47.5% females). For each participant, we included only stimuli of the opposite gender (i.e., all 89 female faces for men and 20 male faces for women).

@@

In the first phase, participants viewed 109 neutral-expression photographs of faces (random order, display time of 3 s) from the American Multiracial Face Database [AMFD, @chen2021broadening]. After each image, participants rated the face on trustworthiness, familiarity, attractiveness, and beauty using visual analog scales. In the second phase, participants were informed that "about half of the previously seen images were AI-generated". The same faces were presented again in a new random order (same display time), followed by ratings of "reality" (whether they believed the image was fake - left anchor - or real - right anchor). 
<!-- @makowski2025too showed that most stimuli yielded highly variable reality ratings, suggesting that idiosyncratic cognitive processes were at play. -->

@@

We fitted 3 models to predict the reality ratings: a ZOIB model, a Beta-Gate model, and the CHOCO model. For all models and each parameter, the full formula was entered: $Real\sim Sex + (1 | Participant) + (1 | Item)$ (with Sex as the main predictor and participants and items entered as random intercepts). The models were run using *brms* [@burkner2017brms] R package, and analyzed using the *easystats* collection of packages [@makowski2025modelbased; @ludecke2020extracting; @patil2022datawizard]. To maximize the comparability across models We used the default priors (uniform) for all models, and we ran 16 chains of 1400 iterations each on the University of Sussex High-Performance Computing (HPC) cluster.

@@

For the population-level effects, we will consider significant and report (using the median of the posterior distribution) effects for which the 95% Credible Interval (CI) does not include zero [and when the probability of direction *pd* is > \~97\%, @makowski2019indices]. For the individual-level parameters (i.e., the random intercepts of each parameter for each participant and each item), we will first analyze their reliability using the Variance-Over-Uncertainty Ratio index (*D-vour*). This index, implemented in the *performance* package [@ludecke2021performance], is inspired by recent work on mixed models reliability [@williams2021beneath; @rouder2024hierarchical], and corresponds to the normalized ratio of observed variability to uncertainty in random effect estimates, defined as:

@@

Where $\sigma_{between}^2$ is the between-group variability (computed as the SD of the random effect point-estimates) and $\mu_{\text{within}}^2$ is the mean squared uncertainty (SE) in random effect estimates (i.e., the average uncertainty).
We use as *D-vour* = 0.666 as the threshold for moderately reliable random effect estimates, which corresponds to a 2:1 ratio of between-group variance to uncertainty.

@@

```{r}
#| label: fig-three
#| fig-cap: "Top: Model comparison revealed that the CHOCO model was a significantly better fit for the data (raw distribution in purple) compared to a ZOIB or Beta-Gate models, capturing its bimodal distribution. Bottom: the CHOCO model can be set to estimate effect on any of its parameters, such as the overall probability of responding on one side as well as the confidence in both choices. We illustrate this by showing the effect of Sex on the distribution of reality beliefs based on a CHOCO model."
#| apa-twocolumn: true  # A Figure Spanning Two Columns When in Journal Mode
#| out-width: "100%"

@@

The ZOIB model suggested that women had higher mean scores of reality beliefs ($\mu_{~\text{Female}} = 0.20,~95\%~ CI~[0.03, 0.37], ~pd = 98.77\%$), less extreme values ($zoi_{~\text{Female}} = -1.24,~95\%~ CI~[-2.46, -0.07], ~pd = 98.25\%$) but more ones relative to zeros ($coi_{~\text{Female}} = 1.63,~95\%~ CI~[0.70, 2.56], ~pd = 99.97\%$).

The Beta-Gate model similarly suggested that women had higher mean scores of reality beliefs ($\mu_{~\text{Female}} = 0.21,~95\%~ CI~[0.02, 0.40], ~pd = 98.52\%$), less extreme values ($pex_{~\text{Female}} = -1.34,~95\%~ CI~[-2.56, -0.17], ~pd = 98.75\%$), and a greater tendency to answer one relative to zero ($bex_{~\text{Female}} = 1.13,~95\%~ CI~[0.34, 1.91], ~pd = 99.76\%$).

The CHOCO model shows that women had a higher probability *p* of judging faces as real ($P_{~\text{Female}} = 0.45,~95\%~ CI~[0.05, 0.86], ~pd = 98.54\%$), but are not more confident when doing so ($confright_{~\text{Female}} = -0.13,~95\%~ CI~[-0.47, 0.21], ~pd = 77.73\%$). However, they were less confident when answering that an image was AI-generated ($confleft_{~\text{Female}} = -0.53,~95\%~ CI~[-0.89, -0.17], ~pd = 99.84\%$). There were also less likely to produce extreme answers ($pex_{~\text{Female}} = -1.19,~95\%~ CI~[-2.40, 0.01], ~pd = 97.64\%$), but no strong evidence supporting a directional bias was observed ($bex_{~\text{Female}} = 0.48,~95\%~ CI~[-0.12, 1.09], ~pd = 94.32\%$).

@@

```{r}
#| label: fig-four
#| fig-cap: "Top: the participant and item-level estimates for the CHOCO parameters, under the distribution of their point-estimates. Reliable effects are characterised by a higher between-group variability relative (the dispersion of the distribution of point-estimates) to its within-group variability (the average uncertainty of individual estimates represented by the error bar). For example, inter-individual variability in the parameters reflecting the confidence in left and right choices is reliably capture, as opposed to the inter-individual variability in the 'bex' parameter. Bottom: correlation matrix between participant-level estimates from different models and empirical indices (e.g., the raw proportion of responses on the right per participant). The CHOCO parameters are easily interpretable, strongly correlating with their empirical counterparts."
#| apa-twocolumn: true  # A Figure Spanning Two Columns When in Journal Mode
#| out-width: "80%"

@@

The explosion of accessibility of state-of-the-art AI tools has made it effortless to generate realistic images, including human faces that are often indistinguishable from real ones [@miller2023ai; @nightingale2022ai; @bozkir2024can]. These synthetic visuals are flooding the cyberspace across various domains, such as art, advertising and entertainment, to education and information. This technological advancement carries an important potential for misuse, such as in disinformation campaigns, scams (e.g., AI-bots, identity theft), and abuse. The democratization of such technology raises pressing concerns about the value of authenticity and the potential erosion of media trust in our increasingly *post-truth* society.

In this evolving landscape, understanding the cognitive mechanisms that underpin our judgments of reality becomes paramount. Despite the increasing prevalence of digitally altered or AI-generated content, humans still rely on certain heuristics to assess authenticity. One such heuristic might be facial attractiveness. Attractiveness appraisals are known to be automatic and unconscious [@luo2019robust; @hung2016unconscious; @hou2023neural], and carry strong real-life consequences, as demonstrated by the large body of literature on the "beauty premium"^[As an eloquent example, @monk2021beholding reported in a large representative US sample that the magnitude of earnings disparities among white women along the perceived attractiveness continuum exceeds in magnitude the canonical black-white race gap.] [@little2021facial; @gulati2024beautiful;@pandey2021face; @kukkonen2024beauty].

Although its role as a potential modulator of reality beliefs remains under-explored, @makowski2025too found significant associations between participants' realness ratings and facial attractiveness, with a potential sexual dymorphism: male participants judged attractive faces as being likely more real, whereas evidence suggested a milder and quadratic (U-shaped) relationship between attractiveness and reality beliefs for females. These findings offer a complementary perspective to those of @miller2023ai, who reported that participants used attractiveness as a distinguishing cue between real and AI-generated faces. These results highlight a possible bidirectional and context-dependent influence of attractiveness on reality judgments, of which the exact shape needs further investigation.

@@

The first sample includes the same participants as study 1 (N = 141, see above). 
For the second sample (data and preprocessing information available at https://github.com/RealityBending/FakeFace2), 248 participants were initially recruited through academic platforms (SONA and SurveySwap). We removed 15 participants with data suggesting low-effort responding as well as those (N=16) that did not believe in the experimental manipulation and were fully confident that all images were real or fake.
<!-- 8 (3.23%) participants with no variation in their responses and 7 (2.92%) participants with no coherence (r < 0.1) b -->
Similarly to the first sample, we included all hetero and bisexual participants and stimuli from the opposite gender, resulting in a final sample included 189 participants (Mean age = 28.4, SD = 14.0, range: [18, 69]; Sex: 76.2% females). The study was approved by the University of Sussex Ethics Board (ER/ST633/1).

@@

### Beauty

>>>>

---
title: "**Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces**"
shorttitle: "Choice-Confidence (CHOCO) Model"
author:
  - name: Dominique Makowski
    corresponding: true
    orcid: 0000-0001-5375-9967
    email: D.Makowski@sussex.ac.uk
    url: https://realitybending.github.io/
    roles:
      - Conceptualization
      - Data curation
      - Formal Analysis
      - Funding acquisition
      - Investigation
      - Methodology
      - Project administration
      - Resources
      - Software
      - Supervision
      - Validation
      - Visualization
      - Writing – original draft
    affiliations:
      - id: "id1"
        name: "University of Sussex"
        department: "School of Psychology"
        city: "Brighton"
        country: "UK"
        postal-code: "BN1 9RH"
      - name: "University of Sussex"
        department:  "Sussex Centre for Consciousness Science"
  - name: Ana Neves  
    orcid: 0009-0006-0020-7599
    role:
      - Data curation
      - Writing – original draft
      - Writing – review & editing
    affiliations:
      - ref: "id1"
  - name: Andy Field 
    orcid: ""
    role:
      - Writing – review & editing
    affiliations:
      - ref: "id1"
author-note:
  status-changes: 
    affiliation-change: ~
    deceased: ~
  disclosures:
    gratitude: |
      ::: {.callout-note icon=false appearance="simple"}
      This preprint is a non-peer-reviewed work from the [**Reality Bending Lab**](https://realitybending.github.io/).
      ![](https://realitybending.github.io/media/ReBeL_LogoOnly_hu11484441381606756729.png){width=20% fig-align="center"}
      :::
    study-registration: ~
    data-sharing: ~
    related-report: ~
    conflict-of-interest: ~
    financial-support: ~
    authorship-agreements: ~
abstract: |
  As AI-generated content becomes increasingly indistinguishable from real stimuli, understanding how individuals form beliefs about the reality of what they perceive is both theoretically and practically urgent. This study introduces the Choice-Confidence (CHOCO) model as a tool to dissect the cognitive mechanisms underlying decision-making judgments. Conceptualized as a mixture of two Beta distributions, CHOCO simultaneously estimates the probability of selecting one category (e.g., "real" vs. "AI-generated") and the confidence associated with each choice. We apply this model to two datasets (N=141 and N=189) in which participants judged whether faces were real photographs or AI-generated images, finding that facial attractiveness systematically increased the likelihood of being judged as "real", particularly for male participants. We discuss how these findings relate to the broader effect of facial attractiveness, and suggest future directions for both the cognitive and psychometric application of the CHOCO framework.
#   **Significance Statement**. TODO.
keywords: "subjective scales, choice confidence model, reality beliefs, AI-generated faces, attractiveness, simulation monitoring"
# Add more keywords if possible: beauty, Ordered Beta model, Bimodal data, decision-making
numbered-lines: true
floatsintext: true
mask: false
bibliography: bibliography.bib
format:
  apaquarto-pdf:
    documentmode: jou
    keep-tex: true
  apaquarto-docx: default
---

@@

<!-- Problem statement -->
Despite significant advancements in psychological science following the replication crisis [@open2015estimating], its progress is still hindered by its sub-optimal (or inappropriate) usage of statistical tools [@makowski2023we; @cumming2014new; @blanca2018current]. 
A prevalent issue is the continued reliance on linear models that assume normally distributed (Gaussian) data^[More specifically, that the outcome is distributed according to a Normal distribution which parameters are expressed as a linear function of the predictors.] - as this assumption often does not hold true for many types of psychological outcomes.
For instance, reaction times typically exhibit skewed distributions, choices can be represented as binary variables, and count data consist of strictly positive integers. 
Applying models that presume normality and model the "mean" of the outcome variable can lead to misinterpretation and potentially misleading conclusions when applied indiscriminately. It is thus important that psychologists use models that can best describe (or generate) the data they collect, to exploit them fully and bring more nuance and accuracy to their conclusions.

<!-- Bounded data, Beta-models and ZOIB -->
Among the most commonly collected data in psychology are responses on subjective scales, such as Likert-type items or visual analog scales, which exhibit some fundamental properties: these responses are bounded (and can be rescaled to a 0-1 range) and frequently display clustering at the extremes. 
Traditional linear models being ill-suited for such data, researchers have turned to using Beta distributions to model these data (instead of Gaussian), suited for continuous data within the (0,1) interval (i.e., excluding extreme responses). 
To address the frequent occurrence of exact zeros and ones (i.e., extreme values), zero-one inflated beta (ZOIB) models have been developed [@ospina2012general] to accommodate the excess of boundary values by incorporating additional components that model the probabilities of responses at 0 and 1 as a separate, independent process.

@@

The Beta-Gate model is a reparametrized Ordered Beta model [@kubinec2023ordered]^[In the Ordered Beta model, the cutpoints on the log-scale are directly used as parameters, instead of being derived from *pex* and *bex*.] available in the *cogmod* package in R (https://github.com/DominiqueMakowski/cogmod), in which participants' answers on bounded scales are conceptualized as latent responses that can fall past a pair of probabilistic "gates" (or cutpoints) that control whether the response is recorded as an extreme (0 or 1) or as a nuanced, continuous value in between (@fig-one). 
The distance of these gates from the edges of the scale varies based on two interpretable parameters: *pex* (the propensity by which people are likely to answer extreme values), and *bex*, a bias toward the upper extreme (1) versus the lower (0). 
A person's internal response that lies close to the edge might be "caught" by a gate and recorded as an extreme, while others pass through to express a continuous response (Beta-distributed with $\mu$ (*mu*) and $\phi$ (*phi*) as its mean and precision parameters).
The Beta-Gate model is based on the idea that extreme values can emerge not just from a fundamentally different underlying processes - as assumed in ZOIB models - but from a common process governed by thresholds of decisiveness and confidence.

```{r}
#| label: fig-one
#| fig-cap: "The Beta-Gate Distribution is a reparametrized ordered Beta model (Kubinec, 2023) that is governed by 4 parameters. 'Mu' and 'phi' correspond to the mean and precision of the continuous part of the distribution (between 0 and 1), and 'pex' (propensity of extremes) and 'bex' (balance of extremes) indirectly control the proportion of zeros and ones by specifying the location of the \"gates\", past which the latent response process is likely to generate extreme values. Specifically, 'pex' defines the total distance of both gates from the extremes (i.e., the sum length of the two orange arrows), and 'bex' determines the proportion of the right gate distance relative to the left (i.e., the length of the right arrow relative to the left one). In this example, the total distance from the extremes is 'pex' = 0.1, with 40% ('bex' = 0.4) of that distance being on the right (and 60% on the left). The left gate is thus located at 0.6, and the right at 1-0.04 = 0.96."
#| fig-align: "center"
#| apa-twocolumn: false  # A Figure Spanning Two Columns When in Journal Mode
#| out-width: "100%"

@@

$$
\begin{cases}
\mathrm{logistic}\left( \mathrm{logit}(pex \cdot (1 - bex)) - \mathrm{logit}(\mu) \right), & x = 0, \\[6pt]
1 - \mathrm{logistic}\left( \mathrm{logit}(1 - pex \cdot bex) - \mathrm{logit}(\mu) \right), & x = 1, \\[6pt]
1 - P(x = 0) - P(x = 1), & 0 < x < 1. \\
\end{cases}
$$

@@

```{r}
#| label: fig-two
#| fig-cap: "Top: The CHOCO Model uses a mixture of Beta-Gate distributions to model separately the right and left sides of the scale (e.g., a rating of whether a statement was 'Truth' vs. 'Lie'), as well as their relative proportion. In this example, the participants are more likely overall (p = 60%) to select the right side of the scale ('Lie') than the left ('Truth'). They are also more confident in their choice (confright = 0.6 vs. confleft = 0.3). Extreme values (zeros and ones) are governed by the same mechanism as for Beta-Gate models. Below: An illustrative dataset in which participants, in condition A (blue), respond as often on the left and right side of the scale, with low confidence (the confidence is concentrated towards the centre). In condition B (orange), they answer more often on the left (but also with low confidence), but their responses on the right are more extreme. As shown in the top table, a typical linear model could lead the imprudent researcher to misleadingly conclude that \"there is no effect of Condition\" (as a linear regression models changes in the mean, which is actually equal for the two conditions). The CHOCO model correctly reveals that condition B is associated with a lower probability but higher confidence of answers on the right side of the scale. Note that the CHOCO model is currently only available through Bayesian modeling, hence we report for consistency pseudo t- and p-values in the table, which correspond respectively to the coefficients divided by their SE, and the values obtained via the pd-to-p conversion [@makowski2019indices]."
#| fig-align: "center"
#| apa-twocolumn: true  # A Figure Spanning Two Columns When in Journal Mode
#| out-width: "70%"

knitr::include_graphics("./figures/fig2final.png")
```

@@

**Study 1** aims to evaluate the CHOCO model’s ability to better capture subjective scale responses that (potentially) reflect an underlying discrete choice, in comparison to existing models such as the ZOIB and Beta-Gate. Specifically, we will assess whether 1) CHOCO provides improved model fit, 2) yields deeper insights into population-level effects than traditional approaches (sex differences in reality beliefs), and 3) allows for the reliable estimation of interpretable individual-level parameters through random effects. **Study 2** will apply this model to more subtle effects, such as the effect of perceived facial attractiveness on reality judgments, and test the ability to fit alternative data structures (scales with ordinal response options and mid-points). To this end, we analyze data from two separate studies in which participants judged whether a face image was AI-generated ("fake") or a real photograph.

@@

The question of "what is real" has become increasingly relevant in the context of recent technological developments, particularly with the rise of artificial intelligence (AI), capable of mass producing highly realistic and relevant content, such as images, videos or text.
These synthetic productions are flooding the cyberspace across various domains, such as art, advertising and entertainment, to education and information. 
The democratization of such technology carries an important potential for misuse, such as in disinformation campaigns and scams (e.g., AI-bots, identity theft), and abuse [e.g., deepfakes, @viola2023designed], and raises pressing concerns about the value of authenticity and the potential erosion of media trust in our increasingly *post-truth* society [@lewandowsky2017beyond].
Critically, as technology improves, traditional cues for detecting what is fake become less reliable (e.g., visual glitches and artefacts in generated images; formulaic generated text, etc.), leading people to increasingly depend on contextual information and cognitive heuristics to assess authenticity, a process referred to as "simulation monitoring" [@makowski2019phenomenal].

@@

Using the open-access data from @makowski2025too, we included all heterosexual and bisexual (as these two groups did not seem to differ based on preliminary analyses and were thus grouped to maximize power) male and female participants, for a final sample of 141 participants (Mean age = 28.4, SD = 9.0, range: [19, 66]; Sex: 47.5% females). For each participant, we included only stimuli of the opposite sex (i.e., all 89 female faces for males and 20 male faces for females).

@@

In the first phase, participants viewed 109 neutral-expression photographs of faces (random order, display time of 3 s) from the American Multiracial Face Database [AMFD, @chen2021broadening]. After each image, participants rated the face on trustworthiness, familiarity, attractiveness, and beauty using visual analog scales. In the second phase, participants were informed that "about half of the previously seen images were AI-generated" (in actuality, all faces were photographs). The same faces were presented again in a new random order (same display time), followed by ratings of "reality" (whether they believed the image was fake - left anchor - or real - right anchor). 
<!-- @makowski2025too showed that most stimuli yielded highly variable reality ratings, suggesting that idiosyncratic cognitive processes were at play. -->

@@

We fitted 3 models to predict the reality ratings: a ZOIB model, a Beta-Gate model, and the CHOCO model. For all models and each parameter, the full formula was entered: $Real\sim Sex + (1 | Participant) + (1 | Item)$ (with Sex as the main predictor and participants and items entered as random intercepts). The models were run using *brms* [@burkner2017brms] R package, and analyzed using the *easystats* collection of packages [@makowski2025modelbased; @ludecke2020extracting; @patil2022datawizard]. To maximize comparability across models, we used the default priors (uniform) for all models, and we ran 16 chains of 1400 iterations each on the University of Sussex High-Performance Computing (HPC) cluster.

@@

For the population-level effects, we will consider "significant" (noteworthy) and report (using the median of the posterior distribution) effects for which the 95% Credible Interval (CI) does not include zero [and when the probability of direction *pd* is > \~97\%, @makowski2019indices]. For the individual-level parameters (i.e., the random intercepts of each parameter for each participant and each item), we will first analyze their reliability using the Variance-Over-Uncertainty Ratio index (*D-vour*). This index, implemented in the *performance* package [@ludecke2021performance], is inspired by recent work on mixed models reliability [@williams2021beneath; @rouder2024hierarchical], and corresponds to the normalized ratio of observed variability to uncertainty in random effect estimates, defined as:

@@

Where $\sigma_{between}^2$ is the between-group variability (computed as the SD of the random effect point-estimates) and $\mu_{\text{within}}^2$ is the mean squared uncertainty (SE) in random effect estimates (i.e., the average uncertainty).
We use as *D-vour* = 0.666 as the threshold for moderately reliable random effect estimates, which corresponds to a 2:1 ratio of between-group variance to uncertainty. Reliable in this context means that the variability between groups (e.g., participants) is greater than the uncertainty in the estimates, suggesting that the parameter can be used to assess inter-individual differences.

@@

```{r}
#| label: fig-three
#| fig-cap: "Top: Model comparison revealed that the CHOCO model was a significantly better fit for the data (raw distribution in purple) compared to a ZOIB or Beta-Gate models, capturing its bimodal distribution. Bottom: the CHOCO model can be set to estimate effect on any of its parameters, such as the overall probability of responding on one side as well as the confidence in both choices. We illustrate this by showing the effect of Sex on the distribution of reality beliefs based on a CHOCO model."
#| fig-align: "center"
#| apa-twocolumn: true  # A Figure Spanning Two Columns When in Journal Mode
#| out-width: "100%"

@@

The ZOIB model suggested that females had higher mean scores of reality beliefs ($\mu_{~\text{Female}} = 0.20,~95\%~ CI~[0.03, 0.37], ~pd = 98.77\%$), less extreme values ($zoi_{~\text{Female}} = -1.24,~95\%~ CI~[-2.46, -0.07], ~pd = 98.25\%$) but more ones relative to zeros ($coi_{~\text{Female}} = 1.63,~95\%~ CI~[0.70, 2.56], ~pd = 99.97\%$).

The Beta-Gate model similarly suggested that females had higher mean scores of reality beliefs ($\mu_{~\text{Female}} = 0.21,~95\%~ CI~[0.02, 0.40], ~pd = 98.52\%$), less extreme values ($pex_{~\text{Female}} = -1.34,~95\%~ CI~[-2.56, -0.17], ~pd = 98.75\%$), and a greater tendency to answer one relative to zero ($bex_{~\text{Female}} = 1.13,~95\%~ CI~[0.34, 1.91], ~pd = 99.76\%$).

The CHOCO model shows that females had a higher probability *p* of judging faces as real ($P_{~\text{Female}} = 0.45,~95\%~ CI~[0.05, 0.86], ~pd = 98.54\%$), but are not more confident when doing so ($confright_{~\text{Female}} = -0.13,~95\%~ CI~[-0.47, 0.21], ~pd = 77.73\%$). However, they were less confident when answering that an image was AI-generated ($confleft_{~\text{Female}} = -0.53,~95\%~ CI~[-0.89, -0.17], ~pd = 99.84\%$). There were also less likely to produce extreme answers ($pex_{~\text{Female}} = -1.19,~95\%~ CI~[-2.40, 0.01], ~pd = 97.64\%$), but no strong evidence supporting a directional bias was observed ($bex_{~\text{Female}} = 0.48,~95\%~ CI~[-0.12, 1.09], ~pd = 94.32\%$).

@@

```{r}
#| label: fig-four
#| fig-cap: "Top: the participant and item-level estimates for the CHOCO parameters, under the distribution of their point-estimates. Reliable effects are characterised by a higher between-group variability relative (the dispersion of the distribution of point-estimates) to its within-group variability (the average uncertainty of individual estimates represented by the error bar). For example, inter-individual variability in the parameters reflecting the confidence in left and right choices is reliably captured, as opposed to the inter-individual variability in the 'bex' parameter. Bottom: correlation matrix between participant-level estimates from different models and empirical indices (e.g., the raw proportion of responses on the right per participant). The CHOCO parameters are easily interpretable, strongly correlating with their empirical counterparts."
#| fig-align: "center"
#| apa-twocolumn: true  # A Figure Spanning Two Columns When in Journal Mode
#| out-width: "80%"

@@

The explosion of accessibility of state-of-the-art AI tools has made it effortless to generate realistic images, including human faces that are often indistinguishable from real ones [@miller2023ai; @nightingale2022ai; @bozkir2024can].
In this evolving landscape, understanding the cognitive mechanisms that underpin our judgments of reality becomes paramount. Despite the increasing prevalence of digitally altered or AI-generated content, humans still rely on certain heuristics to assess authenticity. One such heuristic might be facial attractiveness. Attractiveness appraisals are known to be automatic and unconscious [@luo2019robust; @hung2016unconscious; @hou2023neural], and carry strong real-life consequences, as demonstrated by the large body of literature on the "beauty premium"^[As an eloquent example, @monk2021beholding reported in a large representative US sample that the magnitude of earnings disparities among white women along the perceived attractiveness continuum exceeds in magnitude the canonical black-white race gap.] [@little2021facial; @gulati2024beautiful;@pandey2021face; @kukkonen2024beauty].

Although the role of attractiveness as a potential modulator of reality beliefs remains under-explored, @makowski2025too found significant associations between participants' realness ratings and facial attractiveness, with a potential sexual dymorphism: male participants judged attractive faces as being likely more real, whereas evidence suggested a milder and quadratic (U-shaped) relationship between attractiveness and reality beliefs for females. These findings offer a complementary perspective to those of @miller2023ai, who reported that participants used attractiveness as a distinguishing cue between real and AI-generated faces. These results highlight a possible bidirectional and context-dependent influence of attractiveness on reality judgments, of which the exact shape needs further investigation.

@@

The first sample includes the same participants as study 1 (N = 141, see above). 
For the second sample (data and preprocessing information available at https://github.com/RealityBending/FakeFace2), 248 participants were initially recruited through academic platforms (SONA and SurveySwap). We removed 15 participants with data suggesting low-effort responding (no variation in their responses indicative of straightlining) as well as those (N=16) that did not believe in the experimental manipulation and were fully confident that all images were real or fake.
<!-- 8 (3.23%) participants with no variation in their responses and 7 (2.92%) participants with no coherence (r < 0.1) b -->
Similarly to the first sample, we included all hetero and bisexual participants and stimuli from the opposite gender, resulting in a final sample included 189 participants (Mean age = 28.4, SD = 14.0, range: [18, 69]; Sex: 76.2% females). The study was approved by the University of Sussex Ethics Board (ER/ST633/1).

@@

<!-- ### Beauty -->
